Galois Characterization of Function Fields of \\ Transcendence Degree $1$ over Number Fields

### 2017-08-05 Master Thesis/synopsis.tex
Reference ZS60 VI, §4 Cor 2 to Thm 4
Phrasing

\begin{proof}
See \cite{EP05} Section 2.3 and Corollary 4.1.4, \cite{ZS60} VI, §3 Theorem 3.
\end{proof}

@@

\begin{lemma}\label{lemma-henselization-specialization}
Let $K$ be a field and $v\in\operatorname{Spv}(K)$ with a fixed henselization $K_v$ of $v$. If $w\in\operatorname{Spv}(K)$ such that $w$ is a specialization of $v$, then there is a henselization $K_w$ of $w$ such that $K_v\subset K_w\subset K_v^\text{t}$.
\end{lemma}

@@

For the second assertion, we assume $\ell^\infty\nmid [k:\mathbb{F}_p]$ and $\dim_{\mathbb{F}_\ell}(\Gamma / \ell\Gamma) = 2$. By \cite{En72} Theorem 20.21, we have the following equation of supernatural numbers:
\[ [L:K] = (\Gamma':\Gamma) \cdot [k':k]\cdot p^d \]
for some $d \leq \infty$, where $\Gamma'$ denotes the value group and $k'$ the residue field of the unique extension of $w$ to $L$. By using this equation and the already proven statement from the first part, it suffices to show the following equivalences:
\begin{enumerate}[(i)]
\item $\ell^\infty \nmid [k':k] \iff \ell^\infty \nmid [k':\mathbb{F}_p]$
\item $\ell^\infty\nmid (\Gamma':\Gamma) \iff \dim_{\mathbb{F}_\ell}(\Gamma'/\ell\Gamma')=2$
\end{enumerate}
(i) follows from $[k':\mathbb{F}_p]=[k':k][k:\mathbb{F}_p]$. For (ii) let $T'$ be the absolute inertia group of $L$. By Theorem \ref{theorem-sylowgroups-inertiagroup}, the $\ell$-Sylow group of $T'$ is isomorphic to $\mathbb{Z}_\ell^{r'}$, where $r'=\dim_{\mathbb{F}_\ell}(\Gamma'/\ell\Gamma')$. Furthermore, we have $(\Gamma':\Gamma) = (T:T')$ and the following exact sequence of $\ell$-Sylow groups:
\[ 0 \longrightarrow \mathbb{Z}_\ell^{r'} \longrightarrow \mathbb{Z}_\ell^{r}\longrightarrow (T/T')_\ell\longrightarrow 1 \]
Now $(T/T')_\ell$ is finite if and only if $r'=r=2$.
\end{proof}

@@

Let $w_1,w_2 \in\operatorname{Spv}_\text{d}^{(2)}(K)$ be two different valuations. Then the following statements are equivalent:
\begin{enumerate}[(i)]
\item The composite $K_{w_1}K_{w_2}$ is not algebraically closed.
\item $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ and $v(w_1) = v(w_2)$.
\end{enumerate}
The implication (i)$\implies$(ii) follows from Lemma \ref{2.2}. Now assume (ii) holds. Then $k(v)$ is separably closed where $v=v(w_1)$. We have $K_{w_1}K_{w_2}\subset K_v^\text{t}$ and hence $K_{w_1}K_{w_2}=K_v^\text{t}$ by Theorem \ref{theorem-absoluteinertia}. The extension of $v$ on $K_v^\text{t}$ has value group $\Gamma_v$, which is finitely generated. However, value groups of valuations on an algebraically closed field are divisible, hence $K_v^\text{t}$ can't be algebraically closed.

@@

\begin{proof}
First we assume that $k/\mathbb{Q}$ is a Galois extension and $X$ has a $k$-rational point $P_0$. We can consider $k'$ as a subfield of $k$ via: 
\[\begin{tikzcd}
k'\ar[r] & k(P'_0)\ar[r, "\varphi_{P_0}"', "\cong"] & k(P_0) \ar[r, "\cong"] & k
\end{tikzcd}\]
For all $P\in X_0$ and $P'=\sigma(P)\in X'_0$ we also get an embedding:
\[ \begin{tikzcd}
k \ar[r] & k(P) \ar[r, "\varphi_P^{-1}"', "\cong"] & k(P')
\end{tikzcd} \]
Let us consider the morphism $f:X'\times_{k'} k\to X'$. For $P'\in X'_0$ we have as topological spaces:
\[ f^{-1}(P') = \operatorname{Spec} k(P')\times_{k'} k \]
Any embedding $k\to \overline{k(P')}$ maps $k$ into $k(P')$ because $k/\mathbb{Q}$ is Galois. Thus we have:
\[ k(P')\otimes_{k'} k\cong \prod_{\operatorname{Hom}_{k'}(k, k(P'))} k(P') \]
Now we can see that $f$ splits completely. By Lemma \ref{4.1} we get $[k:k'] = 1$, hence $[k:\mathbb{Q}]= [k':\mathbb{Q}]$.

>>>>

\begin{proof}
See \cite{EP05} Section 2.3 and Corollary 4.1.4, \cite{ZS60} VI, §3 Theorem 3 and VI, §4 Corollary 2 to Theorem 4.
\end{proof}

@@

\begin{lemma}\label{lemma-henselization-specialization}
Let $K$ be a field and $v\in\operatorname{Spv}(K)$ with a fixed henselization $K_v$ of $v$. If $w\in\operatorname{Spv}(K)$ is a specialization of $v$, then there is a henselization $K_w$ of $w$ such that $K_v\subset K_w\subset K_v^\text{t}$.
\end{lemma}

@@

For the second assertion, we assume $\ell^\infty\nmid [k:\mathbb{F}_p]$ and $r=\dim_{\mathbb{F}_\ell}(\Gamma / \ell\Gamma) = 2$. By \cite{En72} Theorem 20.21, we have the following equation of supernatural numbers:
\[ [L:K] = (\Gamma':\Gamma) \cdot [k':k]\cdot p^d \]
for some $d \leq \infty$, where $\Gamma'$ denotes the value group and $k'$ is the residue field of the unique extension of $w$ to $L$. By using this equation and the already proven assertion from the first part, it suffices to show the following equivalences:
\begin{enumerate}[(i)]
\item $\ell^\infty \nmid [k':k] \iff \ell^\infty \nmid [k':\mathbb{F}_p]$
\item $\ell^\infty\nmid (\Gamma':\Gamma) \iff \dim_{\mathbb{F}_\ell}(\Gamma'/\ell\Gamma')=2$
\end{enumerate}
(i) follows from $[k':\mathbb{F}_p]=[k':k][k:\mathbb{F}_p]$. For (ii) let $T'$ be the absolute inertia group of $L$. By Theorem \ref{theorem-sylowgroups-inertiagroup}, the $\ell$-Sylow group of $T'$ is isomorphic to $\mathbb{Z}_\ell^{r'}$, where $r'=\dim_{\mathbb{F}_\ell}(\Gamma'/\ell\Gamma')$. Furthermore, we have $(\Gamma':\Gamma) = (T:T')$ and the following exact sequence of $\ell$-Sylow groups:
\[ 0 \longrightarrow \mathbb{Z}_\ell^{r'} \longrightarrow \mathbb{Z}_\ell^{r}\longrightarrow (T/T')_\ell\longrightarrow 1 \]
Now $(T/T')_\ell$ is finite if and only if $r'=r=2$.
\end{proof}

@@

Let $w_1,w_2 \in\operatorname{Spv}_\text{d}^{(2)}(K)$ be two different valuations. Then the following statements are equivalent:
\begin{enumerate}[(i)]
\item The composite $K_{w_1}K_{w_2}$ is not algebraically closed.
\item $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ and $v(w_1) = v(w_2)$.
\end{enumerate}
The implication (i)$\implies$(ii) follows from Lemma \ref{2.2}. Now assume (ii) holds. Then $k(v)$ is separably closed where $v=v(w_1)$. We have $K_{w_1}K_{w_2}\subset K_v^\text{t}$ and hence $K_{w_1}K_{w_2}=K_v^\text{t}$ by Theorem \ref{theorem-absoluteinertia}. The extension of $v$ on $K_v^\text{t}$ has value group $\Gamma_v$, which is finitely generated. However, value groups of valuations on an algebraically closed field are divisible, hence $K_v^\text{t}$ can't be algebraically closed and neither can the composite $K_{w_1}K_{w_2}$.

@@

\begin{proof}
First we assume that $k/\mathbb{Q}$ is a Galois extension and $X$ has a $k$-rational point $P_0$. We can consider $k'$ as a subfield of $k$ via: 
\[\begin{tikzcd}
k'\ar[r] & k(P'_0)\ar[r, "\varphi_{P_0}"', "\cong"] & k(P_0) \ar[r, "\cong"] & k
\end{tikzcd}\]
For all $P\in X_0$ and $P'=\sigma(P)\in X'_0$ we also get an embedding:
\[ \begin{tikzcd}
k \ar[r] & k(P) \ar[r, "\varphi_P^{-1}"', "\cong"] & k(P')
\end{tikzcd} \]
Let us consider the morphism $f:X'\times_{k'} k\to X'$. For $P'\in X'_0$ we have:
\[ f^{-1}(P') = \operatorname{Spec} k(P')\times_{k'} k \]
Any embedding $k\to \overline{k(P')}$ maps $k$ into $k(P')$ because $k/\mathbb{Q}$ is Galois. Thus we have:
\[ k(P')\otimes_{k'} k\cong \prod_{\operatorname{Hom}_{k'}(k, k(P'))} k(P') \]
Now we can see that $f$ splits completely. By Lemma \ref{4.1} we get $[k:k'] = 1$, hence $[k:\mathbb{Q}]= [k':\mathbb{Q}]$.
